Efficiently Handling Guards in the Low Level Design of RPython's Tracing JIT

### 2012-08-10 talk/vmil2012/paper.tex
some improvements to the evaluation section

From the mentioned benchmarks we collected different datasets to evaluate the
Frequency, the overhead and overall behaviour of guards, the results are
summarized in the remainder of this section. We want to point out three
aspects of guards in particular
\begin{itemize}
  \item Guards are very common operations in traces.
  \item There is overhead associated with guards.
  \item Guard failures are local and rare.
\end{itemize}

All figures in this section do not take garbage collection into account. Pieces
of machine code can be globally invalidated or just become cold again. In both
cases the generated machine code and the related data is garbage collected. The
figures show the total amount of operations that are evaluated by the JIT and
the total amount of code and data that is generated from the optimized traces.

@@

Figure~\ref{fig:benchmarks} summarizes the total number of operations that were
recorded during tracing for each of the benchmarks and what percentage of these
operations are guards. The number of operations was counted on the unoptimized
and optimized traces. The Figure shows that the overall optimization rate for
operations which is between 69.4\% and 83.89\% of the traced operations and the
optimization rate of guards, which is between 65.8\% and 86.2\% of the
operations, are very similar, as could be assumed based on
Figure~\ref{fig:guard_percent}. This indicates that the optimizer can remove
most of the guards, but after the optimization pass guards still account for
15.2\% to 20.2\% of the operations being compiled and later executed, the
frequency of this operation makes it important to store the associated
information efficiently and also to make sure that guard checks are executed
fast.

@@

The overhead that is incurred by the JIT to manage the \texttt{resume data},
the \texttt{low-level resume data} as well as the generated machine code is
shown in Figure~\ref{fig:backend_data}. It shows the total memory consumption
of the code and of the data generated by the machine code backend and an
approximation of the size of the \texttt{resume data} structures for the
different benchmarks mentioned above. The machine code taken into account is
composed of the compiled operations, the trampolines generated for the guards
and a set of support functions that are generated when the JIT starts and which
are shared by all compiled traces. The size of the \texttt{low-level resume
data} is the size of the compressed mapping from registers and stack to
IR-level variables and finally the size of the \texttt{resume data} is an
approximation of the size of the compressed high-level resume data as described
in Section~\ref{sec:Resume Data}\todo{explain why it is an approximation}.

@@

Why the efficient storing of the \texttt{resume data} is a central concern in the design
of guards is illustrated by Figure~\ref{fig:backend_data}, this Figure shows
the size of the compressed \texttt{resume data}, the approximated size of
storing the \texttt{resume data} without compression and the size of
compressing the data to calculate the size of the resume data using the
\texttt{xz} compression tool, which is a ``general-purpose data compression
software with high compression ratio'' used to approximate the best possible
compression for the \texttt{resume data}.\footnote{\url{http://tukaani.org/xz/}}.

The results show that the current approach of compression and data sharing only
requires 18.3\% to 31.1\% of the space compared to the naive approach. This
shows that large parts of the resume data are redundant and can be stored more
efficiently through using the techniques described above. On the other hand
comparing the results to the xz compression which only requires between 17.1\%
and 21.1\% of the space required by our compression shows that the compression
is not optimal but a trade-off between the required space and the time needed
to build a good compressed representation of the compressed resume data for the
large amount of guards present in the traces.

\subsection{Guard Failures}
\label{sub:guard_failure}
The last point in this discussion is the frequency of guard failures.
Figure~\ref{fig:failing_guards} presents for each benchmark a list of the
relative amounts of guards that ever fail and of guards that fail more than 200
times. For guards that fail more than 200 times, as described before, a trace
is recorded that starts from the guard, patching the guard so that later
failures execute the new trace instead of taking the side-exit. Hence the
numbers presented for guards that fail more than 200 times represent the 200
failures up to the compilation of the bridge and all executions of the then
attached bridge.

@@

\todo{add a footnote about why guards have a threshold of 200}

>>>>

From the mentioned benchmarks we collected different datasets to evaluate the
frequency, the overhead and overall behaviour of guards, the results are
summarized in the remainder of this section. We want to point out three
aspects of guards in particular
\begin{itemize}
  \item Guards are very common operations in traces.
  \item There is overhead associated with guards.
  \item Guard failures are local and rare.
\end{itemize}

All figures in this section do not take garbage collection of machine code into account. Pieces
of machine code can be globally invalidated or just become cold again. In both
cases the generated machine code and the related data is garbage collected. The
figures show the total amount of operations that are evaluated by the JIT and
the total amount of code and data that is generated from the optimized traces.

@@

Figure~\ref{fig:benchmarks} summarizes the total number of operations that were
recorded during tracing for each of the benchmarks and what percentage of these
operations are guards. The number of operations was counted on the unoptimized
and optimized traces. The Figure shows that the overall optimization rate for
operations which is between 69.4\% and 83.89\% of the traced operations and the
optimization rate of guards, which is between 65.8\% and 86.2\% of the
operations, are very similar, as could be assumed based on
Figure~\ref{fig:guard_percent}. This indicates that the optimizer can remove
most of the guards, but after the optimization pass guards still account for
15.2\% to 20.2\% of the operations being compiled and later executed.
The frequency of guard operations makes it important to store the associated
information efficiently and also to make sure that guard checks are executed
quickly.

@@

The overhead that is incurred by the JIT to manage the \texttt{resume data},
the \texttt{low-level resume data} as well as the generated machine code is
shown in Figure~\ref{fig:backend_data}. It shows the total memory consumption
of the code and of the data generated by the machine code backend and an
approximation of the size of the \texttt{resume data} structures for the
different benchmarks mentioned above. The machine code taken into account is
composed of the compiled operations, the trampolines generated for the guards
and a set of support functions that are generated when the JIT starts and which
are shared by all compiled traces. The size of the \texttt{low-level resume
data} is the size of the compressed mapping from registers and stack to
IR-level variables and finally the size of the \texttt{resume data} is an
approximation of the size of the compressed high-level resume data as described
in Section~\ref{sec:Resume Data}.\footnote{
The size of the resume data is not measured at runtime, but reconstructed from
log files.}

@@

Why the efficient storing of the \texttt{resume data} is a central concern in the design
of guards is illustrated by Figure~\ref{fig:backend_data}. This figure shows
the size of the compressed \texttt{resume data}, the approximated size of
storing the \texttt{resume data} without compression and
an approximation of the best possible compression of the resume data by
compressing the data using the
\texttt{xz} compression tool, which is a ``general-purpose data compression
software with high compression ratio''.\footnote{\url{http://tukaani.org/xz/}}

The results show that the current approach of compression and data sharing only
requires 18.3\% to 31.1\% of the space compared to a naive approach. This
shows that large parts of the resume data are redundant and can be stored more
efficiently through using the techniques described above. On the other hand
comparing the results to the xz compression which only requires between 17.1\%
and 21.1\% of the space required by our compression shows that the compression
is not optimal but a trade-off between the required space and the time needed
to build a good compressed representation of the compressed resume data for the
large amount of guards present in the traces.

\subsection{Guard Failures}
\label{sub:guard_failure}
The last point in this discussion is the frequency of guard failures.
Figure~\ref{fig:failing_guards} presents for each benchmark a list of the
relative amounts of guards that ever fail and of guards that fail more than 200
times.\footnote{
    The threshold of 200 is rather high. It was picked experimentally to give
    good results for long-running programs.
}
As described before, for guards that fail more than 200 times, a trace
is recorded that starts from the guard. Afterwards the guard is patched so that later
failures execute the new trace instead of taking the side-exit. Hence the
numbers presented for guards that fail more than 200 times represent the 200
failures up to the compilation of the bridge and all executions of the then
attached bridge.

@@
